---
title: "reach.utils: capabilities tour"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{reach.utils: capabilities tour}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

```{r library}
library(reach.utils)
```

## Overview

reach.utils is the shared utility layer for the Flode ecosystem: date and
time handling, water-year calendars, time series quality checks,
configuration loading, a pipeline activity runner, structured logging, path
helpers, argument assertions, and project scaffolding. Every other
`reach.*` package calls down into this one rather than reimplementing these
pieces itself.

This vignette walks through each module with runnable examples. The project
scaffolding functions at the end write files and open editors, so those
chunks are shown but not evaluated.

## Date and time

Operational systems run in UTC; local time must never enter a pipeline.
`as_utc()` and `parse_datetime()` get you there from whatever format the
data arrived in.

```{r datetime-parse}
as_utc("2024-01-15 06:00:00")
as_utc("15/01/2024 06:00", fmt = "%d/%m/%Y %H:%M")

# parse_datetime() tries a prioritised list of formats commonly seen in
# EA hydrometric data, so you don't have to know which one you've got
parse_datetime(c("2024-01-15T06:00:00Z", "15/01/2024 06:00", "2024-01-15"))
```

`floor_to()` and `snap_to_datetime()` floor or round a timestamp to a grid
unit; `seq_datetime()` generates a regular series.

```{r datetime-grid}
x <- as_utc("2024-01-15 06:37:22")
floor_to(x, "15 minutes")
snap_to_datetime(x, "15 minutes")
snap_to_datetime(x, "1 hour")

readings <- seq_datetime("2024-01-01", "2024-01-02", by = "15 mins")
length(readings)
```

`water_year()` labels a timestamp by the EA water year it falls in (starts
1 October, labelled by the year it begins).

```{r datetime-water-year}
water_year(as_utc(c("2024-09-30", "2024-10-01")))
```

`detect_gaps()` and `is_complete_series()` are the standard pre-processing
checks before anything downstream trusts a series to be regular.

```{r datetime-gaps}
# A 15-minute series missing two steps around 01:00
# (always include seconds in a timestamp string: as_utc()'s auto-parser can
# misread a bare "HH:MM" scalar, even though it reads the same string fine
# as part of a longer vector)
gappy <- as_utc(c(
  "2024-01-01 00:00:00", "2024-01-01 00:15:00",
  "2024-01-01 01:00:00", "2024-01-01 01:15:00"
))
detect_gaps(gappy, by = "15 mins")

is_complete_series(
  as_utc(c("2024-01-01 00:00:00", "2024-01-01 00:15:00", "2024-01-01 00:30:00")),
  from = "2024-01-01 00:00:00", to = "2024-01-01 00:30:00", by = "15 mins"
)
```

`format_duration()` turns a number of seconds into a readable string, used
throughout the logging helpers below.

```{r datetime-duration}
format_duration(8070)
format_duration(90)
format_duration(0.4)
```

## Water-year calendar

The `epoch` functions build on `water_year()` for calendar-aware iteration
and reporting.

```{r epoch-bounds}
water_year_bounds(2023)
water_year_seq(2020, 2023)
```

`split_water_years()` partitions a series (and, optionally, parallel
values) into one element per water year present.

```{r epoch-split}
dates <- as_utc(c("2023-06-01", "2023-11-01", "2024-03-01", "2024-11-01"))
split_water_years(dates)
split_water_years(dates, values = c(1.1, 2.2, 3.3, 4.4))
```

`complete_water_years()` reports which water years in a series have full
15-minute (or other interval) coverage, and `label_season()` assigns a
seasonal label under three different schemes.

```{r epoch-complete-season}
two_years <- seq_datetime("2022-10-01", "2024-09-30 23:45:00", by = "15 mins")
complete_water_years(two_years)

sample_dates <- as_utc(c("2024-01-15", "2024-04-15", "2024-07-15", "2024-10-15"))
label_season(sample_dates)
label_season(sample_dates, scheme = "meteorological")
label_season(sample_dates, scheme = "hydrological")
```

## Quality control

Six `check_*()` functions each flag one failure mode in a time series;
`qc_series()` runs a configurable subset of them together and returns one
named list of results.

```{r qc-individual}
qc_times  <- seq_datetime("2024-01-01", "2024-01-01 01:00:00", "15 mins")
qc_values <- c(1.0, 1.1, 1.1, 1.1, NA, 99.9)

check_duplicates(qc_times)
check_monotonic(qc_times)
check_bounds(qc_times, qc_values, min = 0, max = 10)
check_flatline(qc_times, qc_values, n = 3L)
check_rate_of_change(qc_times, qc_values, max_change = 5)
check_na_runs(qc_times, qc_values, max_run = 1L)
```

Each returns zero rows when the check passes, so they compose cleanly into
a pipeline assertion. `qc_series()` runs several at once:

```{r qc-series}
qc_series(
  qc_times, qc_values,
  bounds = c(0, 10), max_change = 5, flatline_n = 3
)
```

## Configuration

`load_config()` reads a YAML or JSON file into a named list; the rest of
the `config` functions query, validate, merge, and type-coerce it.

```{r config-load}
cfg_path <- tempfile(fileext = ".yml")
writeLines(c(
  "site:",
  "  id: \"4206TH\"",
  "database:",
  "  host: localhost",
  "  port: 5432",
  "timeout: \"30\"",
  "debug: \"true\""
), cfg_path)

cfg <- load_config(cfg_path)
get_config_val(cfg, c("database", "host"))
get_config_val(cfg, "missing_key", default = "fallback")

validate_config(cfg, c("site", "database"))
```

`config_val_as()` coerces a value pulled from YAML text to the type you
actually need, including the `"true"`/`"false"` and date quirks YAML
parsers tend to leave as strings.

```{r config-coerce}
config_val_as(cfg, "timeout", "integer")
config_val_as(cfg, "debug", "logical")
```

`merge_configs()` layers an override config (environment-specific paths,
say) over a base one, recursing into nested lists so only the differing
keys need to be specified:

```{r config-merge}
base    <- list(db = list(host = "localhost", port = 5432L), timeout = 30L)
overlay <- list(db = list(host = "prod-server"))
merge_configs(base, overlay)
```

`config_from_env()` builds a config list from environment variables under
a prefix, and `expand_config_paths()` resolves any relative-looking path
values in a config against a base directory:

```{r config-env-paths}
Sys.setenv(REACH__DB__HOST = "prod-server", REACH__TIMEOUT = "60")
config_from_env("REACH")
Sys.unsetenv(c("REACH__DB__HOST", "REACH__TIMEOUT"))

expand_config_paths(
  list(input = "data/flow.csv", host = "localhost"),
  base_dir = "/srv/pipeline"
)
```

## Pipeline activities

`register_activity()` lets any package expose its functions as named
activities; `run_pipeline()` then drives a sequence of them from a single
YAML file. This is how a multi-step, multi-package pipeline becomes a
config file rather than a bespoke script.

```{r activities-register}
ingest_demo <- function(n_readings = 96L, site_id) {
  log_info("Ingesting {n_readings} readings for site {site_id}")
  start <- as_utc("2024-01-01")
  end   <- start + (n_readings - 1L) * 900  # 900s = 15 min
  data.frame(
    datetime = seq_datetime(start, end, by = "15 mins"),
    flow_cms = round(pmax(0, 10 + cumsum(rnorm(n_readings, sd = 0.3))), 3)
  )
}

qc_demo <- function(max_val) {
  x <- seq_datetime("2024-01-01", "2024-01-01 01:00:00", "15 mins")
  v <- c(1.0, 1.1, 1.1, 1.1, NA, 99.9)
  qc_series(x, v, bounds = c(0, max_val), max_change = 5)
}

register_activity("ingest", ingest_demo, package = "reach.utils")
register_activity("qc_check", qc_demo, package = "reach.utils")
list_activities()
```

The config supplies a `global` block shared by every activity (here,
`site_id` and `max_val`), filled in automatically for any activity
argument left unset in its own `args`:

```{r activities-config}
pipeline_path <- tempfile(fileext = ".yml")
writeLines(c(
  "global:",
  "  site_id: \"4206TH\"",
  "  max_val: 50",
  "",
  "activities:",
  "  - name: ingest",
  "    package: reach.utils",
  "    args:",
  "      n_readings: 8",
  "  - name: qc_check",
  "    package: reach.utils"
), pipeline_path)

cat(readLines(pipeline_path), sep = "\n")
```

```{r activities-run}
set.seed(1)
results <- run_pipeline(pipeline_path)
names(results)
results[["reach.utils::ingest"]]
results[["reach.utils::qc_check"]]
```

`run_pipeline()` resolves every activity before running any of them, so a
typo in the config aborts before a partial run:

```{r activities-missing, error = TRUE}
bad_path <- tempfile(fileext = ".yml")
writeLines(c("activities:", "  - name: does_not_exist", "    package: reach.utils"), bad_path)
run_pipeline(bad_path)
```

## Logging

`run_pipeline()` logs through these helpers; they are equally useful in a
hand-written script. Messages interpolate `{expressions}` from wherever
they were called, using [cli](https://cli.r-lib.org/) formatting.

```{r logging-basic}
site_id <- "4206TH"
n_rows  <- 96L
log_info("Processing station {site_id}: {n_rows} records")
```

```{r logging-warn}
gap_n <- 3L
log_warn("Gap detected in station {site_id}: {gap_n} missing steps")
```

`log_error()` logs and raises an error, so pipeline code can rely on it to
stop execution:

```{r logging-error, error = TRUE}
log_error("Failed to load NWP grid: {.path missing_file.nc}")
```

`log_debug()` only emits when the `reach.utils.log_level` option is set to
`"debug"`, for diagnostic detail that should stay quiet in normal runs:

```{r logging-debug}
log_debug("This is suppressed by default")

options(reach.utils.log_level = "debug")
log_debug("Processing {.val 42} rows")
options(reach.utils.log_level = NULL)
```

`log_timed()` wraps an expression, times it, and logs the elapsed duration
via [format_duration()]; `log_progress()` and `log_section()` round out
console output for long-running loops:

```{r logging-timed}
result <- log_timed(Sys.sleep(0.05), label = "short pause")

for (i in seq_len(3)) log_progress(i, 3, label = "site ABCD")

log_section("Backfill")
```

`log_to_file()` mirrors a plain-text, timestamped line to a file instead of
(or alongside) the console, for runs where console output isn't captured:

```{r logging-file}
log_path <- tempfile(fileext = ".log")
log_to_file(log_path, "Backfill completed for site ABCD")
cat(readLines(log_path), sep = "\n")
```

## Paths

A small set of path helpers for the things every pipeline script needs:
create a directory if absent, fail loudly if an input is missing, resolve
a path without requiring it to exist yet, find the newest file matching a
pattern, and swap an extension.

```{r paths}
out_dir <- ensure_dir(file.path(tempdir(), "reach-utils-demo", "outputs"))
dir.exists(out_dir)

demo_file <- file.path(out_dir, "results.csv")
writeLines("a,b\n1,2", demo_file)
check_file_exists(demo_file)

resolve_path(tempdir(), "reach-utils-demo", "..", "outputs")

swap_ext("data/flow.csv", "parquet")
swap_ext("outputs/run.tar.gz", "zip")
```

```{r paths-find}
Sys.sleep(0.01)  # ensure a distinct mtime from demo_file
newest_file <- file.path(out_dir, "results_v2.csv")
writeLines("a,b\n3,4", newest_file)
find_latest_file(out_dir, "*.csv")
```

`check_file_exists()` aborts with a clear message when the file is absent,
rather than letting a cryptic error surface further downstream:

```{r paths-missing, error = TRUE}
check_file_exists(file.path(out_dir, "does_not_exist.csv"))
```

## Assertions

Guard a function's entry point rather than letting a type or shape
mistake propagate into a confusing failure three calls later.

```{r assert-pass}
ts     <- as_utc(c("2024-01-01", "2024-01-02"))
values <- c(1.2, 3.4)

assert_posixct(ts)
assert_numeric(values)
assert_scalar(42)
assert_length(c(0, 1), n = 2L)
assert_same_length(ts, values, args = c("timestamps", "values"))
assert_choice("linear", c("linear", "locf", "constant"))
```

Each one aborts with an informative, argument-named message on failure:

```{r assert-fail, error = TRUE}
assert_numeric("not a number")
```

```{r assert-fail-choice, error = TRUE}
assert_choice("cubic", c("linear", "locf", "constant"))
```

## Project scaffolding

The `templates` functions scaffold a new project, a config file, a script
or function stub, a Quarto README or report, and an RStudio new-script
template. These write to disk (and `create_script_template()` and
`manual_edit` touch the user's RStudio configuration or open an editor),
so this section is illustrative rather than run.

```{r templates, eval = FALSE}
# Scaffold a full project: R/, data/raw/, data/processed/, config/,
# outputs/, logs/, tests/testthat/, .gitignore, and a starter config
create_project(author = "Forecasting and Warning Team")

# The starter config written above, or standalone:
create_config(site_id = "42001")

# A new script pre-filled with the standard header
create_script(
  file_name = "data_import",
  file_path = "R",
  author    = "Forecasting and Warning Team",
  email     = "forecasting@environment-agency.gov.uk"
)

# A function stub with a roxygen2 skeleton
create_function("aggregate_flow", parameters = c("x", "values", "by"))

# Set the RStudio new-script template to the reach.utils standard header
create_script_template()

# Quarto README and analytical report skeletons
create_readme(format = "github", author = "Forecasting and Warning Team",
             readme_title = "My Reach Project")
create_report(
  format       = "html",
  file_name    = "flood_analysis_2024",
  report_title = "Flood Frequency Analysis",
  author       = "Forecasting and Warning Team"
)
```

## Where this fits

reach.utils has no dependency on any other `reach.*` package; everything
above runs standalone. Packages that do depend on it, such as reach.io,
call these same functions rather than reimplementing date handling,
config loading, or logging. See that package's vignettes for activities
registered against real pipeline functions (`run_backfill()`,
`run_incremental()`) rather than the toy ones used here.
